---
title: "bikesharing"
author: "Maciej"
date: "2023-08-17"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

## Scenario

You are a junior data analyst working in the marketing analyst team at
Cyclistic, a bike-share company in Chicago. The director of marketing
believes the company's future success depends on maximizing the number
of annual memberships. Therefore, your team wants to understand how
casual riders and annual members use Cyclistic bikes differently. From
these insights, your team will design a new marketing strategy to
convert casual riders into annual members. But first, Cyclistic
executives must approve your recommendations, so they must be backed up
with compelling data insights and professional data visualizations.

### Three questions will guide the future marketing program:

1.  How do annual members and casual riders use Cyclistic bikes
    differently?
2.  Why would casual riders buy Cyclistic annual memberships?
3.  How can Cyclistic use digital media to influence casual riders to
    become members?

# Bike-share project

The objectives of this project is:

-   Grab data from given source : [Link for
    files](https://divvy-tripdata.s3.amazonaws.com/index.html)
-   Consolidate/aggregate data of bike trips from bike-sharing company
-   Make analysis due to business task.
-   Share insights and recommendations

## Installing required packages

```{r loading packages}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(dplyr)
library(here)
library(aws.s3)

```

## Setting global language

```{r setting language}

Sys.setlocale("LC_TIME", "C")

```

## Setting enviroment for aws s3

```{r}

Sys.setenv("AWS_DEFAULT_REGION" = "eu-north-1")
```

## Checking connection to bucket

```{r}

bucket_exists(
  bucket = "s3://analitycal-projects-data/", 
  region = "eu-north-1"
)
```

## Listing bucket contents and saving file names 

```{r}

bucket_objects <- get_bucket("analitycal-projects-data")
first_12_objects <- head(bucket_objects,12)
names_of_s3_files <- sapply(first_12_objects, function(obj) obj$Key)

```

## Downloading csv files from bucket

There is 12 CSV files to download in cyclist_data/csv_files folder.
Downloaded files are stored at project directory

```{r, eval = FALSE}

subfolder = "cyclist_data/csv_files/"
files_date <- c("2022_08", "2022_09", "2022_10", "2022_11", "2022_12", "2023_01", "2023_02", "2023_03", "2023_04", "2023_05", "2023_06", "2023_07")
file_ending = "_trip_data.csv"



for (name_of_s3_file in names_of_s3_files) {
  
  save_object(
    object = name_of_s3_file,
    bucket = "s3://analitycal-projects-data/", 
    region = "eu-north-1",
    file = name_of_s3_file
  )
 
}


```

## Setting relative path to csv files

```{r}

main_path <- here("cyclist_data", "csv_files")
print(main_path)

```

## Reading CSV files and saving to list

Reading already downloaded files that are stored at cyclist_data/csv_files

```{r reading csv files}

list_of_month_trips <- list()
i <- 1

for (name_of_s3_file in names_of_s3_files) {
  file_path <- name_of_s3_file
  data <- read.csv(file_path)
  list_of_month_trips[[i]] <- data
  i <- i + 1
}

```

## Dropping irrelevant columns

Columns associated with place of trips are irrelevant because business tasks are related
to when and how questions but not where. For example information about the most frequent start/end station will not help to find patterns that can help to find a way to attract customers to buy a membership. Eventually it can be useful for placing advertisments in these places.

```{r}

relevant_columns <- c("ride_id","started_at","ended_at","member_casual")
list_length = length(list_of_month_trips)
for(i in 1:list_length)
{
  list_of_month_trips[[i]] <- subset(list_of_month_trips[[i]],select=relevant_columns)
  
}

```

## Consolidating the 12 datasets into big one df

```{r binding rows}

all_trips <- bind_rows(list_of_month_trips)

```

## Summary

```{r summary}

summary(all_trips)
head(all_trips)

```

## Adding date columns

```{r adding date columns}

all_trips$date <- as.Date(all_trips$started_at) #The default format is yyyy-mm-dd
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")

```

## Changing column types

Making changes to column type started_at and ended_at from char to
posixct due to latter computation - ride_length

```{r changing type of columns}

all_trips$started_at <-as.POSIXct(all_trips$started_at,
                                             format = "%Y-%m-%d %H:%M:%S")
all_trips$ended_at <-as.POSIXct(all_trips$ended_at,
                                             format = "%Y-%m-%d %H:%M:%S")


```

## Adding ride_length column

Adding ride_length column that compute the difference in seconds between
start and end of the trip

```{r adding ride_length column}

all_trips$ride_length <- as.numeric(all_trips$ended_at - all_trips$started_at)

```

## Filtering and making new subset of records

Deleting rows that have ride length less than 0 - these rows are
irrelevant from analysis point of view and rows that have empty string
in member_casual

```{r filtering and creating new subset}

all_trips_v2 <- all_trips %>%
  filter(ride_length > 0, member_casual %in% c("member", "casual"))

```

## Comparing members and casual riders

Aggregating groups the unique values in member_casual column and compute
given statistical value for each group

```{r aggregating and compute}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = summary)
```

## Defining the order of caterogical values

Making order of day_of_week column and their values without changing the
factual order of records

```{r making order}

all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

```

## Aggregating the average ride time

Aggregating by each day and each member type in ordered manner

```{r aggregating}

aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)

```

## Plot 1

Comparing the number of rides by each type of membership

```{r plotting 1}

data <- all_trips_v2 %>%
  group_by(member_casual) %>%
  summarise(number_of_rides = n()) %>%
  ungroup() %>%
  mutate(percent_change = ((number_of_rides - lag(number_of_rides)) / lag(number_of_rides)) * 100)

p <- ggplot(data, aes(x = member_casual, y = number_of_rides, fill = member_casual)) +
  geom_col(width=0.6) +
  geom_text(aes(label = paste0(round(percent_change, 1), "%")), 
      vjust = -0.2) +
  labs(title = "Number of rides with percentage difference",
       y = "Number of rides",
       x = "Type of membership",
       caption ="Regular users use bikes more (amount of rentals) by 71.2 on average.",
       fill="Type of membership") +
  theme(plot.caption = element_text(margin=margin(t=20), size=12,hjust=0))

print(p)

```

## Plot 2

Comparing the average duration of rides by each type of membership

```{r plotting 2}

data <- all_trips_v2 %>%
  group_by(member_casual) %>%
  summarise(average_duration = mean(ride_length)) %>%
  ungroup() %>%
  mutate(percent_change = ((average_duration - lag(average_duration)) / lag(average_duration)) * 100)

p <- ggplot(data, aes(x = member_casual, y = average_duration, fill = member_casual)) +
  geom_col(width=0.6) +
  geom_text(aes(label = paste0(round(percent_change, 1), "%")), 
             vjust = -0.2) +
  labs(title = "Average duration of rides with percentage difference",
       y = "Ride duration",
       x = "Type of membership",
       caption ="On the other hand, ordinary users ride longer on average by 53.1% more",
       fill="Type of membership") +
  theme(plot.caption = element_text(margin=margin(t=20), size=12,hjust=0))
print(p)
  
```

## Creating and saving comparision of number of rides and average duration grouped by type of member and weekday

```{r creating comparision}

rides_and_mean_by_day_and_member <- all_trips_v2 %>% 
  mutate(weekday = day_of_week) %>%
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n()
  ,average_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday)	

```

## Plot 3

Comparing number of rides within each day of week by member type

```{r plotting 3}

rides_and_mean_by_day_and_member  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(title = "Number of rides by each day of weekday",
       y = "Number of rides",
       x = "Day of week",caption ="Casual users use bike sharing most often on weekends.\nRegular use most often at the week.",
       fill="Type of membership") +
  theme(plot.caption = element_text(margin=margin(t=20), size=12,hjust=0))

```

## Plot 4

Comparing mean of rides within each day of week by member type

```{r plotting 4}

rides_and_mean_by_day_and_member %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(title = "Average duration of rides by each day of weekday",
       y = "Ride duration",
       x = "Day of week",
       caption = "Casual users use bike longer on weekends than on weekdays.\nRegular users and their length of time are stable",
       fill="Type of membership") +
  theme(plot.caption = element_text(margin=margin(t=20), size=12,hjust=0))

```

## Plot 5

Comparing number of rides within each month by member type

```{r plotting 5}

all_trips_v2 %>% 
  group_by(member_casual, month) %>%
  summarise(number_of_rides = n() 
  ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, month) %>%
  ggplot(aes(x = month, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
    labs(title = "Number of rides by each month",
       y = "Number of rides",
       x = "Month",
       caption = "In both cases, it can be seen that the number of trips increases as the temperature increases",
       fill="Type of membership") +
  theme(plot.caption = element_text(margin=margin(t=20), size=12,hjust=0))

```

## Plot 6

Comparing number of rides within each month by member type

```{r plotting 6}

all_trips_v2 %>% 
  group_by(member_casual, month) %>%
  summarise(number_of_rides = n()							
  ,average_duration = mean(ride_length)) %>% 	
  arrange(member_casual, month) %>%
  ggplot(aes(x = month, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") +
    labs(title = "Average duration of rides by each month",
       y = "Ride duration",
       x = "Month",
       caption = "In both cases it can be seen that the duration of trips is bigger when it is warm",
       fill="Type of membership") +
  theme(plot.caption = element_text(margin=margin(t=20), size=12,hjust=0))

```

## Questions and answers


#### Why would casual riders buy Cyclistic annual memberships ?

-   To save on rides because they use them longer than the users who have annual memberships
-   It may also be about convenience of use
-   They may be encouraged by advertising campaigns

#### How can Cyclistic use digital media to influence casual riders to become members?

-   The idea is to pay attention to the length of the rides and the high cost associated with it - on weekends, casual users use bike-sharing longer and more frequent. It is worth to pay attention to possibility of saving when they buy annual membership. They could take advantage by longer trips and less cost associated with this longer trips. The point is to make them more flexible on weekends and charge them less costs.
-   It can be seen thar regular customers use it most often during the week. It is possible they use this solution to get to work. Maybe some of casual users also commute by bike-sharing. The idea is to encourage casual users to start commuting by bike-sharing or use it more frequent. This could led to change of thinking of casual riders to buy annual membership.
-   Finally, recommendation that is based on increasing investment outlay before warm months and keep promoting during these months 
