---
title: "Bike sharing"
author: "Maciej"
date: "2023-08-17"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

## Scenario

You are a junior data analyst working in the marketing analyst team at
Cyclistic, a bike-share company in Chicago. The director of marketing
believes the company's future success depends on maximizing the number
of annual memberships. Therefore, your team wants to understand how
casual riders and annual members use Cyclistic bikes differently. From
these insights, your team will design a new marketing strategy to
convert casual riders into annual members. But first, Cyclistic
executives must approve your recommendations, so they must be backed up
with compelling data insights and professional data visualizations.

### Three questions will guide the future marketing program:

1.  How do annual members and casual riders use Cyclistic bikes
    differently?
2.  Why would casual riders buy Cyclistic annual memberships?
3.  How can Cyclistic use digital media to influence casual riders to
    become members?

# Bike-share project

The objectives of this project is:

-   Grab data from given source : [Link for
    files](https://divvy-tripdata.s3.amazonaws.com/index.html)
-   Consolidate/aggregate data of bike trips from bike-sharing company
-   Make analysis due to business task.
-   Share insights and recommendations

## Installing required packages

```{r loading packages}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(dplyr)
library(here)
library(aws.s3)

```

## Setting global language

```{r setting language}

Sys.setlocale("LC_TIME", "C")

```

## Setting enviroment for aws s3

```{r}
# this is a big NO NO, sharing keys in public repositories can expose company data
# Recommended fix: https://docs.oracle.com/en/database/oracle/machine-learning/oml4r/1.5.1/oread/creating-and-modifying-environment-variables-on-windows.html#GUID-DD6F9982-60D5-48F6-8270-A27EC53807D0 
# something like above, every person that needs to access the data should have personal KEYs 
# if you don't mind sharing data with others, consider changing s3 bucket accessibility to public.
# ofc for learning this is a good thing you used s3 and learned some cool things
# depending on the permissions on those keys I recommend you revoke them in AWS UI
Sys.setenv("AWS_ACCESS_KEY_ID" = "AKIATCI2ENCEIBNJIINP",
           "AWS_SECRET_ACCESS_KEY" = "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",
           "AWS_DEFAULT_REGION" = "eu-north-1"
           )
```

## Checking connection to bucket

```{r}
bucket_exists(
  bucket = "s3://analitycal-projects-data/", 
  region = "eu-north-1"
)
```

## Listing bucket contents

```{r}

get_bucket("analitycal-projects-data")

```

## Downloading csv files from bucket

There is 12 CSV files to download in cyclist_data/csv_files folder.
Downloaded files are stored at project directory

```{r, eval = FALSE}


# nice approach to extend your knowledge you can try to find aws function that will list all files names in the bucket
# try looking here: https://search.r-project.org/CRAN/refmans/aws.s3/html/get_bucket.html

subfolder = "cyclist_data/csv_files/"
files_date <- c("2022_08", "2022_09", "2022_10", "2022_11", "2022_12", "2023_01", "2023_02", "2023_03", "2023_04", "2023_05", "2023_06", "2023_07")
file_ending = "_trip_data.csv"


# for better code redability you can try to indent content of code blocks like below:
for (file_date in files_date) {
  file_path_to_download = paste("cyclist_data/csv_files/", file_date, file_ending, sep="")

  file_path_to_save = paste("data/csv_files/", file_date, file_ending, sep="")

  save_object(
    object = file_path_to_download,
    bucket = "s3://analitycal-projects-data/", 
    region = "eu-north-1",
    file = file_path_to_save
  )
 
}


```

## Setting relative path to csv files

```{r}
main_path <- here("data", "csv_files")
print(main_path)
```

## Reading CSV files and saving to list

Reading already downloaded files that are stored at data/csv_files

```{r reading csv files}

# you already have those values in vector defined
# try to reuse variables this brings less confision to the code reader 
months <- c("2022_08", "2022_09", "2022_10", "2022_11", "2022_12", "2023_01", "2023_02", "2023_03", "2023_04", "2023_05", "2023_06", "2023_07")
list_of_month_trips <- list()

# nice readable for loop below! 
for (month in months) {
  file_path <- paste(main_path, paste(month, "_trip_data.csv", sep = ""), sep = "/")
  data <- read.csv(file_path)
  list_of_month_trips[[month]] <- data
}
```

## Dropping irrelevant columns

```{r}

# a brief explanation on why those columns are irrelevant could be usefull 
relevant_columns <- c("ride_id","started_at","ended_at","member_casual")
list_length = length(list_of_month_trips)
for(i in 1:list_length)
{
  list_of_month_trips[[i]] <- subset(list_of_month_trips[[i]],select=relevant_columns)
  
}

# This can be achieved by:
  

```

## Consolidating the 12 datasets into big one df

```{r binding rows}
all_trips <- bind_rows(list_of_month_trips)

```

## Summary

```{r summary}
summary(all_trips)
head(all_trips)
```

## Adding date columns

```{r adding date columns}
all_trips$date <- as.Date(all_trips$started_at) #The default format is yyyy-mm-dd
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")

```

## Changing column types

Making changes to column type started_at and ended_at from char to
posixct due to latter computation - ride_length

```{r changing type of columns}
all_trips$started_at <-as.POSIXct(all_trips$started_at,
                                             format = "%Y-%m-%d %H:%M:%S")
all_trips$ended_at <-as.POSIXct(all_trips$ended_at,
                                             format = "%Y-%m-%d %H:%M:%S")


```

## Adding ride_length column

Adding ride_length column that compute the difference in seconds between
start and end of the trip

```{r adding ride_length column}
all_trips$ride_length <- as.numeric(all_trips$ended_at - all_trips$started_at)

```

## Filtering and making new subset of records

Deleting rows that have ride length less than 0 - these rows are
irrelevant from analysis point of view and rows that have empty string
in member_casual

```{r filtering and creating new subset}



all_trips_v2 <- all_trips %>%
  filter(ride_length > 0, member_casual %in% c("member", "casual"))




```

## Comparing members and casual riders

Aggregating groups the unique values in member_casual column and compute
given statistical value for each group

```{r aggregating and compute}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = summary)
```

## Defining the order of caterogical values

Making order of day_of_week column and their values without changing the
factual order of records

```{r making order}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

## Aggregating the average ride time

Aggregating by each day and each member type in ordered manner

```{r aggregating}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

## Plot 1

Comparing the number of rides by each type of membership

```{r plotting 1}
data <- all_trips_v2 %>%
  group_by(member_casual) %>%
  summarise(number_of_rides = n()) %>%
  ungroup() %>%
  mutate(percent_change = ((number_of_rides - lag(number_of_rides)) / lag(number_of_rides)) * 100)

p <- ggplot(data, aes(x = member_casual, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  geom_text(aes(label = paste0(round(percent_change, 1), "%")), 
            position = position_dodge(width = 0.9), vjust = -0.5) +
  labs(title = "Number of rides with percentage difference",
       y = "Number of rides",
       x = "Type of membership")
print(p)
  
```

## Plot 2

Comparing the average duration of rides by each type of membership

```{r plotting 2}
data <- all_trips_v2 %>%
  group_by(member_casual) %>%
  summarise(average_duration = mean(ride_length)) %>%
  ungroup() %>%
  mutate(percent_change = ((average_duration - lag(average_duration)) / lag(average_duration)) * 100)

p <- ggplot(data, aes(x = member_casual, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") +
  geom_text(aes(label = paste0(round(percent_change, 1), "%")), 
            position = position_dodge(width = 0.9), vjust = -0.5) +
  labs(title = "Average duration of rides with percentage difference",
       y = "Average duration of rides",
       x = "Type of membership")
print(p)
  
```

## Creating and saving comparision of number of rides and average duration grouped by type of member and weekday

```{r creating comparision}
rides_and_mean_by_day_and_member <- all_trips_v2 %>% 
  mutate(weekday = day_of_week) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday) %>%  #groups by usertype and weekday
  summarise(number_of_rides = n()							#calculates the number of rides and average duration 
  ,average_duration = mean(ride_length)) %>% 		# calculates the average duration
  arrange(member_casual, weekday)	

```

## Plot 3

Comparing number of rides within each day of week by member type

```{r plotting 3}
rides_and_mean_by_day_and_member  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(title = "Number of rides by each day of weekday  ",
       y = "Number of rides",
       x = "Day of week")

```

## Plot 4

Comparing mean of rides within each day of week by member type

```{r plotting 4}
rides_and_mean_by_day_and_member %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(title = "Mean of rides by each day of weekday  ",
       y = "Average duration of rides",
       x = "Day of week")

```

## Plot 5

Comparing number of rides within each month by member type

```{r plotting 5}
all_trips_v2 %>% 
  group_by(member_casual, month) %>%
  summarise(number_of_rides = n() 
  ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, month) %>%
  ggplot(aes(x = month, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
    labs(title = "Number of rides by each month",
       y = "Number of rides",
       x = "Month")
```

## Plot 6

Comparing number of rides within each month by member type

```{r plotting 6}
all_trips_v2 %>% 
  group_by(member_casual, month) %>%
  summarise(number_of_rides = n()							
  ,average_duration = mean(ride_length)) %>% 	
  arrange(member_casual, month) %>%
  ggplot(aes(x = month, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") +
    labs(title = "Mean of rides by each month",
       y = "Average duration of rides",
       x = "Month")
```

## Questions and answers

#### How do annual members and casual riders use Cyclistic bikes differently?

> asnwers should be given in english keep whole project in the same language.
> intepretation could be placed below plots it belogns to

-   Głównie częściej korzystają stali użytkownicy (średnio o 63,8%
    więcej ), ale dłuzej korzystają zwykli użytkownicy (średnio o 56%
    więcej)
-   Zwykli korzystają najczęściej w weekendy, stali korzystają
    najczęściej w tygodniu. W sobote i w niedziele liczba przejazdów sa
    prawie takie same
-   W ciagu miesiąca od 1 do 31 średni czas przejazdu jest stały dla
    stałych użytkowników, a dla zwykłych jest bardziej zróżnicowany
-   W ciagu roku w obydwu przypadkach widać, że liczba przejzdów wzrasta
    gdy tempteratura wzrasta
-   To samo tyczy się długości przejazdów, są one nieco dłuzsze gdy jest
    cieplej

#### Why would casual riders buy Cyclistic annual memberships ?

-   By zaoszczędzić na przejazdach, ponieważ korzystają z nich dłużej
-   Również może chodzić o wygodę w korzystaniu
-   Mogą być zachęceni przez kampanie reklamowe

#### How can Cyclistic use digital media to influence casual riders to become members?

-   Pomysł na zasadzie, zwrócenia uwagi na długośc przejazdów i wysoki
    koszt z tym związany - w weekendy zwykli użytkownicy częściej i
    dłużej korzystają z bike-sharingu. Warto zwrócić uwagę w reklamie by
    skorzystali oni z rocznego by jescze bardziej skorzystać z usługi
    oraz zaoszczędzić pieniądze i cieszyć się przy weekndowych
    przejażdzkach.
-   Widać, że stali klienci korzystają najczęściej w tygodniu i możliwe,
    że korzystają z tego rozwiązania by dojechać do pracy. Możliwe, że
    jest jakaś część klientów okazjonalnych, która też dojeżdża do pracy
    ale nie wykupiła rocznego abonamentu, albo wybiera bike-sharing co
    jakiś czas przy dojazdach. Warto skupić się na tym by zachęcić ich
    do częstszych dojazdów do pracy oraz wykupu rocznego abonamentu.
-   Rekomendacja, która opiera się na zwiększeniu nakładów
    inwestycyjnych, tuż przed okresem z zwiększoną temperaturą lub w
    trakcie.
